---
title: "Club Classification Models"
author: "Devin Basley"
date: "2024-03-28"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(caret)
library(xgboost)
library(purrr)
library(randomForest)
```



# Load Data

```{r}

# Load in Trackman Golf Data
tman_data <- read.csv("~/Downloads/SquareFace/2024-cmu/Data Cleaning/Output Data/Cleaned_All_Clubs.csv")



tman_data <- tman_data %>%
  mutate(player_id = case_when(Player == "Alexis Sudjianto" ~ 1,
                               Player == "Daniel Rodgers" ~ 2,
                               Player == "Jay Saluja" ~ 3,
                               Player == "Ryan Dannegger" ~ 4,
                               Player == "Stephen Dai" ~ 5,
                               Player == "Andrew" ~ 6,
                               Player == "David Zhang" ~ 7,
                               Player == "Justin Chan" ~ 8,
                               Player == "Samantha Wang" ~ 9,
                               Player == "Steven Shea" ~ 10,
                               Player == "Ashley Liu" ~ 11,
                               Player == "Folker" ~ 12,
                               Player == "Justin Luan" ~ 13,
                               Player == "Sense sangkagoon" ~ 14,
                               Player == "Bli" ~ 15,
                               Player == "Hank" ~ 16,
                               Player == "Kaylin Yeoh" ~ 17,
                               Player == "Simulator Player" ~ 18,
                               Player == "Brian Xu" ~ 19,
                               Player == "Jack Sonne" ~ 20,
                               Player == "Nikita Jadhav" ~ 21,
                               Player == "Sriya" ~ 22))

unknown_clubs_df <- tman_data %>%
  filter(Club == "Unknown Club")


unknown_clubs <- unknown_clubs_df %>%
  select(c(Max.Height...Dist., Max.Height...Height, Ball.Speed, Club.Speed, 
           Attack.Angle, Dynamic.Lie, Dyn..Loft, player_id, Swing.Plane, Club))

  



st_club <- c("4 Iron", "5 Iron", "6 Iron", "7 Iron", "8 Iron", "9 Iron",
             "Pitching Wedge", "Driver")

tman_data <- tman_data %>%
  filter(Club %in% st_club) %>%
  mutate(club_id = case_when(Club == "Driver" ~ 0,
                               Club == "Pitching Wedge" ~ 1,
                               Club == "4 Iron" ~ 2,
                               Club == "5 Iron" ~ 3,
                               Club == "6 Iron" ~ 4,
                               Club == "7 Iron" ~ 5,
                               Club == "8 Iron" ~ 6,
                               Club == "9 Iron" ~ 7))

tman_data <- tman_data %>%
  select(c(club_id, Max.Height...Dist., Max.Height...Height, Ball.Speed, Club.Speed, 
           Attack.Angle, Dynamic.Lie, Dyn..Loft, player_id, Swing.Plane))



```

# xgBoost Model CV

```{r}

trainSize <- round(0.80 * nrow(tman_data))

set.seed(1234) 
trainIndex <- sample(nrow(tman_data), trainSize)
trainDF <- tman_data %>% dplyr::slice(trainIndex)
testDF <- tman_data %>% dplyr::slice(-trainIndex)
dtrain <- xgb.DMatrix(data = as.matrix(select(trainDF, -club_id)), label = trainDF$club_id)
dtest <- xgb.DMatrix(data = as.matrix(select(testDF, -club_id)), label = testDF$club_id)



set.seed(524)
club_CV <- xgb.cv(params = list(eta = 0.3),
               data = dtrain, 
               num_class = length(unique(trainDF$club_id)),
               booster = "gbtree",  
               objective = "multi:softmax", 
               nfold = 10,
               nrounds = 500,
               early_stopping_rounds = 10,
               print_every_n = 5)




```



# Assess xgBoost Performance

```{r}
# Determine the optimal number of boosting rounds
optimal_round <- club_CV$best_iteration



# Print optimal number of boosting rounds
cat("Optimal Number of Boosting Rounds:", optimal_round, "\n")

# Assess model performance
best_model <- xgb.train(params = list(eta = 0.3),
                        data = dtrain,
                        num_class = length(unique(trainDF$club_id)),
                        booster = "gbtree",
                        objective = "multi:softmax",
                        nrounds = optimal_round)

# Make predictions on the test set
pred <- predict(best_model, dtest)

# Evaluate predictions
accuracy <- mean(pred == testDF$club_id)
cat("Accuracy on Test Set:", accuracy, "\n")


```



# Random Forest Model Cross Validation

```{r}

trainSize <- round(0.80 * nrow(tman_data))

set.seed(1234) 
trainIndex <- sample(nrow(tman_data), trainSize)
trainDF <- tman_data %>% dplyr::slice(trainIndex)
testDF <- tman_data %>% dplyr::slice(-trainIndex)




 # rf_model <- randomForest(as.factor(club_id) ~ `Max.Height...Dist.` + `Max.Height...Height` + 
 #                            Ball.Speed + Club.Speed + Attack.Angle +
 #                            Dynamic.Lie + Dyn..Loft + player_id , 
 #                          data = trainDF, ntree = 500)

 #varImp(rf_model)

set.seed(123)
# Define training control
ctrl <- trainControl(method = "cv",    # Cross-validation method
                     number = 10)       # Number of folds

set.seed(123)
# Train the model using cross-validation
rf_model_cv <- train(as.factor(club_id) ~ `Max.Height...Dist.` + `Max.Height...Height` + 
                       Ball.Speed + Club.Speed + Attack.Angle +
                       Dynamic.Lie + Dyn..Loft + player_id , 
                     data = trainDF, 
                     method = "rf",           # Random forest method
                     trControl = ctrl,        # Use defined training control
                     ntree = 500)

# Print cross-validation results
#print(rf_model_cv)


rf_preds <- predict(rf_model_cv, newdata = testDF)



# Evaluate predictions
rf_accuracy <- mean(rf_preds == testDF$club_id)
cat("Random Forest Accuracy on Test Set:", rf_accuracy, "\n")

```

# Predict on Unknown Club

```{r}

unknown_preds <- predict(rf_model_cv, newdata = unknown_clubs)

table(unknown_preds)

unknown_clubs_df$pred_club_id <- unknown_preds

```


# Unknown Club Data Cleaning

```{r}

unknown_clubs_df <- unknown_clubs_df %>% 
  mutate(Club = case_when(pred_club_id == 0 ~ "Driver",
                          pred_club_id == 1 ~ "Pitching Wedge",
                          pred_club_id == 2 ~ "4 Iron",
                          pred_club_id == 3 ~ "5 Iron",
                          pred_club_id == 4 ~ "6 Iron",
                          pred_club_id == 5 ~ "7 Iron",
                          pred_club_id == 6 ~ "8 Iron",
                          pred_club_id == 7 ~ "9 Iron"))

# The next few lines limit the data via our "good shot" criteria
unknown_clubs_df <- unknown_clubs_df %>%
  filter(!(Club == "Driver" & `Est..Total.Flat...Side` > 15))

unknown_clubs_df <- unknown_clubs_df %>%
  filter(!(Club == "Driver" & `Est..Total.Flat...Side` < -15))

unknown_clubs_df <- unknown_clubs_df %>%
  filter(!(Club == "9 Iron" & `Est..Total.Flat...Side` > 5))

unknown_clubs_df <- unknown_clubs_df %>%
  filter(!(Club == "9 Iron" & `Est..Total.Flat...Side` < -5))

unknown_clubs_df <- unknown_clubs_df %>%
  filter(!(Club == "8 Iron" & `Est..Total.Flat...Side` > 5))

unknown_clubs_df <- unknown_clubs_df %>%
  filter(!(Club == "8 Iron" & `Est..Total.Flat...Side` < -5))

unknown_clubs_df <- unknown_clubs_df %>%
  filter(!(Club == "7 Iron" & `Est..Total.Flat...Side` > 7.5))

unknown_clubs_df <- unknown_clubs_df %>%
  filter(!(Club == "7 Iron" & `Est..Total.Flat...Side` < -7.5))

unknown_clubs_df <- unknown_clubs_df %>%
  filter(!(Club == "6 Iron" & `Est..Total.Flat...Side` > 7.5))

unknown_clubs_df <- unknown_clubs_df %>%
  filter(!(Club == "6 Iron" & `Est..Total.Flat...Side` < -7.5))

unknown_clubs_df <- unknown_clubs_df %>%
  filter(!(Club == "5 Iron" & `Est..Total.Flat...Side` > 10))

unknown_clubs_df <- unknown_clubs_df %>%
  filter(!(Club == "5 Iron" & `Est..Total.Flat...Side` < -10))

unknown_clubs_df <- unknown_clubs_df %>%
  filter(!(Club == "4 Iron" & `Est..Total.Flat...Side` > 10))

unknown_clubs_df <- unknown_clubs_df %>%
  filter(!(Club == "4 Iron" & `Est..Total.Flat...Side` < -10))

unknown_clubs_df <- unknown_clubs_df %>%
  filter(!(Club == "Pitching Wedge" & `Est..Total.Flat...Side` > 5))

unknown_clubs_df <- unknown_clubs_df %>%
  filter(!(Club == "Pitching Wedge" & `Est..Total.Flat...Side` < -5))

# Filter out shots that do not go far
unknown_clubs_df <- filter(unknown_clubs_df, !(
    Club == "Driver" & `Est..Total.Flat...Length` < quantile(
        filter(
            unknown_clubs_df, Club == "Driver"
        )$`Est..Total.Flat...Length`, 0.15
    )
))

unknown_clubs_df <- filter(unknown_clubs_df, !(
    Club == "9 Iron" & `Est..Total.Flat...Length` < quantile(
        filter(
            unknown_clubs_df, Club == "9 Iron"
        )$`Est..Total.Flat...Length`, 0.10
    )
))

unknown_clubs_df <- filter(unknown_clubs_df, !(
    Club == "8 Iron" &`Est..Total.Flat...Length` < quantile(
        filter(
            unknown_clubs_df, Club == "8 Iron"
        )$`Est..Total.Flat...Length`, 0.10
    )
))

unknown_clubs_df <- filter(unknown_clubs_df, !(
    Club == "7 Iron" & `Est..Total.Flat...Length` < quantile(
        filter(
            unknown_clubs_df, Club == "7 Iron"
        )$`Est..Total.Flat...Length`, 0.10
    )
))

unknown_clubs_df <- filter(unknown_clubs_df, !(
    Club == "6 Iron" & `Est..Total.Flat...Length` < quantile(
        filter(
            unknown_clubs_df, Club == "6 Iron"
        )$`Est..Total.Flat...Length`, 0.10
    )
))

unknown_clubs_df <- filter(unknown_clubs_df, !(
    Club == "5 Iron" & `Est..Total.Flat...Length` < quantile(
        filter(
            unknown_clubs_df, Club == "5 Iron"
        )$`Est..Total.Flat...Length`, 0.10
    )
))

unknown_clubs_df <- filter(unknown_clubs_df, !(
    Club == "4 Iron" & `Est..Total.Flat...Length` < quantile(
        filter(
            unknown_clubs_df, Club == "4 Iron"
        )$`Est..Total.Flat...Length`, 0.10
    )
))

unknown_clubs_df <- filter(unknown_clubs_df, !(
    Club == "Pitching Wedge" & `Est..Total.Flat...Length` < quantile(
        filter(
            unknown_clubs_df, Club == "Pitching Wedge"
        )$`Est..Total.Flat...Length`, 0.10
    )
))

#write_csv(df_st_club, "Output Data/Cleaned_Standard_Clubs.csv")








# Calculate shaft_lean for known clubs
unknown_clubs_df <- unknown_clubs_df %>%
  mutate(shaft_lean = ifelse(Club == "Driver", 10.5 - `Dyn..Loft`,
    ifelse(Club == "4 Iron", 22 - `Dyn..Loft`,
      ifelse(Club == "5 Iron", 25 - `Dyn..Loft`,
        ifelse(Club == "6 Iron", 29 - `Dyn..Loft`,
          ifelse(Club == "7 Iron", 33 - `Dyn..Loft`,
            ifelse(Club == "8 Iron", 37 - `Dyn..Loft`,
              ifelse(Club == "9 Iron", 41 - `Dyn..Loft`,
                ifelse(Club == "Pitching Wedge", 45 - `Dyn..Loft`,
                  NA
                )
              )
            )
          )
        )
      )
    )
  ))


```


# Unknown Club - Driver Side GAM

```{r}

library(mgcv)

pred_driver_df <- unknown_clubs_df %>%
  filter(Club == "Driver")

attach(pred_driver_df)

pred_driver_side <- gam(`Max.Height...Side` ~ s(Club.Speed) + s(Attack.Angle) +
                           s(Swing.Plane) + s(Face.Angle) + s(Club.Path) +
                           s(shaft_lean), 
                         data = pred_driver_df)

summary(pred_driver_side)
print(pred_driver_side)




```


# Unknown Club - Driver Height GAM

```{r}
pred_driver_height <- gam(`Max.Height...Height` ~ s(Club.Speed) + s(Attack.Angle) +
                           s(Swing.Plane) + s(Face.Angle) + s(Club.Path) +
                           s(shaft_lean), 
                         data = pred_driver_df)
summary(pred_driver_height)
print(pred_driver_height)


```

# Split Predicted Driver Data into Quartiles

```{r}

# SPLIT THE DATA INTO QUARTERS
# First distance quartile
q1_dist <- quantile(pred_driver_df$`Est..Total.Flat...Length`, 0.25)
# Second Distance quartile
median_dist <- quantile(pred_driver_df$`Est..Total.Flat...Length`, 0.50)
# Third Distance quartile
q3_dist <- quantile(pred_driver_df$`Est..Total.Flat...Length`, 0.75)

# 0-25% of data range to classify a straight
straight_range1 <- 8 / 2 # Distance from zero (target)
# 25-50% of data range to classify a straight
straight_range2 <- 10 / 2 # Distance from zero (target)
# 50-75% of data range to classify a straight
straight_range3 <- 12 / 2 # Distance from zero (target)
# 75-100% of data range to classify a straight
straight_range4 <- 14 / 2 # Distance from zero (target)

# 0-25% of data, divide height standard deviation by this
sd_divisor1 <- 2
# 25-50% of data, divide height standard deviation by this
sd_divisor2 <- 2
# 0-25% of data, divide height standard deviation by this
sd_divisor3 <- 3
# 25-100% of data, divide height standard deviation by this
sd_divisor4 <- 3

```


# Predicted Driver Side Refitting

```{r}

# Get side predictions from model and stor them in a column
pred_driver_df$prediction_side <- predict(pred_driver_side, pred_driver_df)

# Initially split for drivers and fades based on sign
pred_driver_df$side_class <- ifelse(pred_driver_df$prediction_side > 0, "Draw", "Fade")

# Loop to classify striaghts
for (i in 1:nrow(pred_driver_df)) {
  # See if point is in the first quartile of shot distances
  if (pred_driver_df[i, "Est..Total.Flat...Length"] < q1_dist) {
    # See if the shot meets straight criteria for q1 distance
    if (pred_driver_df[i, "prediction_side"] < straight_range1 &
      pred_driver_df[i, "prediction_side"] > -straight_range1) {
      # Classify the point as a straight
      pred_driver_df[i, "side_class"] <- "Straight"
    }
    # See if point is in the second quartile of shot distances
  } else if (pred_driver_df[i, "Est..Total.Flat...Length"] < median_dist) {
    # See if the shot meets straight criteria for q2 distance
    if (pred_driver_df[i, "prediction_side"] < straight_range2 &
      pred_driver_df[i, "prediction_side"] > -straight_range2) {
      # Classify the point as a straight
      pred_driver_df[i, "side_class"] <- "Straight"
    }
    # See if point is in the third quartile of shot distances
  } else if (pred_driver_df[i, "Est..Total.Flat...Length"] < q3_dist) {
    # See if the shot meets straight criteria for q3 distance
    if (pred_driver_df[i, "prediction_side"] < straight_range3 &
      pred_driver_df[i, "prediction_side"] > -straight_range3) {
      # Classify the point as a straight
      pred_driver_df[i, "side_class"] <- "Straight"
    }
    # See if point is in the fourth quartile of shot distances
  } else {
    # See if the shot meets straight criteria for q4 distance
    if (pred_driver_df[i, "prediction_side"] < straight_range4 &
      pred_driver_df[i, "prediction_side"] > -straight_range4) {
      # Classify the point as a straight
      pred_driver_df[i, "side_class"] <- "Straight"
    }
  }
}


```


# Predicted Driver Height Refit

```{r}

# Get height predictions from model
pred_driver_df$prediction_height <- predict(pred_driver_height, pred_driver_df)
# Define height class variable with all initially Medium
pred_driver_df$height_class <- "Medium"

# Limit data to first quartile
pred_driver_df_height1 <- filter(pred_driver_df, `Est..Total.Flat...Length` < q1_dist)
# Get mean and standard deviation
mean_lim_height1 <- mean(pred_driver_df_height1$prediction_height)
sd_lim_height1 <- sd(pred_driver_df_height1$prediction_height)

# Limit data to second quartile
pred_driver_df_height2 <- filter(pred_driver_df, `Est..Total.Flat...Length` >= q1_dist)
pred_driver_df_height2 <- filter(pred_driver_df_height2, `Est..Total.Flat...Length` < median_dist)
# Get mean and standard deviation
mean_lim_height2 <- mean(pred_driver_df_height2$prediction_height)
sd_lim_height2 <- sd(pred_driver_df_height2$prediction_height)

# Limit data to third quartile
pred_driver_df_height3 <- filter(pred_driver_df, `Est..Total.Flat...Length` >= median_dist)
pred_driver_df_height3 <- filter(pred_driver_df_height3, `Est..Total.Flat...Length` < q3_dist)
mean_lim_height3 <- mean(pred_driver_df_height3$prediction_height)
sd_lim_height3 <- sd(pred_driver_df_height3$prediction_height)

# Limit data to fourth quartile
pred_driver_df_height4 <- filter(pred_driver_df, `Est..Total.Flat...Length` >= q3_dist)
# Get mean and standard deviation
mean_lim_height4 <- mean(pred_driver_df_height4$prediction_height)
sd_lim_height4 <- sd(pred_driver_df_height4$prediction_height)

# For loop to classify varying heights
for (i in 1:nrow(pred_driver_df)) {
  # Check to see if distance is in first quartile
  if (pred_driver_df[i, "Est..Total.Flat...Length"] < q1_dist) {
    # Check to see if shot is high
    if (pred_driver_df[i, "prediction_height"] > (mean_lim_height1 + sd_lim_height1 / sd_divisor1)) {
      pred_driver_df[i, "height_class"] <- "High"
    }
    # Check to see if shot is low
    if (pred_driver_df[i, "prediction_height"] < (mean_lim_height1 - sd_lim_height1 / sd_divisor1)) {
      pred_driver_df[i, "height_class"] <- "Low"
    }
    # Check to see if distance is in second quartile
  } else if (pred_driver_df[i, "Est..Total.Flat...Length"] < median_dist) {
    # Check to see if shot is high
    if (pred_driver_df[i, "prediction_height"] > (mean_lim_height2 + sd_lim_height2 / sd_divisor2)) {
      pred_driver_df[i, "height_class"] <- "High"
    }
    # Check to see if shot is low
    if (pred_driver_df[i, "prediction_height"] < (mean_lim_height2 - sd_lim_height2 / sd_divisor2)) {
      pred_driver_df[i, "height_class"] <- "Low"
    }
    # Check to see if distance is in first quartile
  } else if (pred_driver_df[i, "Est..Total.Flat...Length"] < q3_dist) {
    # Check to see if shot is high
    if (pred_driver_df[i, "prediction_height"] > (mean_lim_height3 + sd_lim_height3 / sd_divisor3)) {
      pred_driver_df[i, "height_class"] <- "High"
    }
    # Check to see if shot is low
    if (pred_driver_df[i, "prediction_height"] < (mean_lim_height3 - sd_lim_height3 / sd_divisor3)) {
      pred_driver_df[i, "height_class"] <- "Low"
    }
    # Check to see if distance is in first quartile
  } else {
    # Check to see if shot is high
    if (pred_driver_df[i, "prediction_height"] > (mean_lim_height4 + sd_lim_height4 / sd_divisor4)) {
      pred_driver_df[i, "height_class"] <- "High"
    }
    # Check to see if shot is low
    if (pred_driver_df[i, "prediction_height"] < (mean_lim_height4 - sd_lim_height4 / sd_divisor4)) {
      pred_driver_df[i, "height_class"] <- "Low"
    }
  }
}

```


# Distributions

```{r}

# Choose the shot shape you want to investigate
shot_shape_df <- filter(pred_driver_df, side_class == "Straight" & height_class == "Medium")

par(mfrow= c(2,3))

# Create a histogram for all parameters of a chosen shot shape
hist(shot_shape_df$`Club.Speed`, breaks = 10, main = "Club Speed")
abline(v = median(shot_shape_df$`Club.Speed`))
hist(shot_shape_df$`Attack.Angle`, breaks = 10, main = "Attack Angle")
abline(v = median(shot_shape_df$`Attack.Angle`))
hist(shot_shape_df$`Swing.Plane`, breaks = 10, main = "Swing Plane")
abline(v = median(shot_shape_df$`Swing.Plane`))
hist(shot_shape_df$`Face.Angle`, breaks = 10, main = "Face Angle")
abline(v = median(shot_shape_df$`Face.Angle`))
hist(shot_shape_df$`Club.Path`, breaks = 10, main = "Club Path")
abline(v = median(shot_shape_df$`Club.Path`))
hist(shot_shape_df$shaft_lean, breaks = 10, main = "Shaft Lean")
abline(v = median(shot_shape_df$shaft_lean))

```



# Extract Predicted Driver Parameters

```{r}

# Create a data frame template to keep ideal parameters
df_ideal_driver <- data.frame(Club=character(), Height_Class=character(), Side_Class=character(),
                       Club_Path_Angle=double(), Lie_Angle=double(), Attack_Angle=double(),
                       Club_Speed=double(), Shaft_Lean=double(), Face_Angle=double())

# Lists for shot shapes
heights = c("High", "Medium", "Low")
sides = c("Draw", "Straight", "Fade")

# Variable for row iterator
row = 1

# Loop though each shot shape combination
for (height in heights){
    for (side in sides){
        # Filter to only specific shot shape
        shot_shape_df <- filter(pred_driver_df, side_class == side & height_class == height)
        
        # Store club and shot shape
        df_ideal_driver[row, "Club"] <- "Driver"
        df_ideal_driver[row, "Height_Class"] <- height
        df_ideal_driver[row, "Side_Class"] <- side
        
        # Get ideal parameters from medians
        df_ideal_driver[row, "Club_Path_Angle"] <- mean(shot_shape_df$`Club.Path`)
        df_ideal_driver[row, "Lie_Angle"] <- mean(shot_shape_df$`Swing.Plane`)
        df_ideal_driver[row, "Attack_Angle"] <- mean(shot_shape_df$`Attack.Angle`)
        df_ideal_driver[row, "Club_Speed"] <- mean(shot_shape_df$`Club.Speed`)
        df_ideal_driver[row, "Shaft_Lean"] <- mean(shot_shape_df$shaft_lean)
        df_ideal_driver[row, "Face_Angle"] <- mean(shot_shape_df$`Face.Angle`)
        
        # Iterate to next row
        row = row + 1
    }
}

# Show data frame with ideal parameters
df_ideal_driver

```



# Extracted Parameter Table

```{r}
library(kableExtra)
# Dummy df for sharing table
df_ideal_driver_sharable <- data.frame(
    Shot_Shape = paste(df_ideal_driver$Height_Class,df_ideal_driver$Side_Class),
    Club_Path_Angle = df_ideal_driver$Club_Path_Angle,
    Lie_Angle = df_ideal_driver$Lie_Angle,
    Attack_Angle = df_ideal_driver$Attack_Angle,
    Club_Speed = df_ideal_driver$Club_Speed,
    Shaft_Lean = df_ideal_driver$Shaft_Lean,
    Face_Angle = df_ideal_driver$Face_Angle)

# Output kable
kable(df_ideal_driver_sharable, escape = TRUE, caption = "Ideal Parameter Estimates for Driver") %>%
  kable_styling(latex_options = "hold_position")


```